The Companion Mass Distribution of extremely low-mass white dwarfs

### 2014-10-16 documents/letter/ms.tex
Added figshare description
update MAP numbers

\begin{abstract}
Measuring the masses of companions to single-line spectroscopic binary stars
is (in general) not possible because of the unknown orbital plane inclination angles. 
Even when the mass of the visible star can be measured, only a lower limit can be placed
on the mass of the unseen companion. However, since the orbital geometry of the systems should be isotropically
distributed, for a large enough, unbiased sample, the 
companion mass distribution can be deconvolved from the distributions of observables.
In this work, we construct a hierarchical probabilistic model to infer properties of unseen 
companion stars given orbital period and projected radial velocity measurements of the primary. We use a mixture of 
two Gaussians to model the true companion mass distribution to represent 
populations of white dwarfs (WD) and neutron stars (NS). We run this model on mock data sets in order to test the
robustness of our inferences. We then apply our model to 55 of the 61 white dwarfs in 
the extremely low mass (ELM) WD survey. We find maximum a posteriori (MAP) parameters for
the WD companion population of $\mu_{\rm WD} = 0.73~\Msun$ and $\sigma_{\rm WD} = 0.26~\Msun$ 
with a NS fraction consistent with zero, $f_{\rm NS} = 0\%$. However, the marginal posterior 
distribution over NS fraction has a significant tail up to $f_{\rm NS} \approx 12\%$.
We make samples from the posterior distribution publicly available and provide a  
script for computing the companion mass distribution and the NS probability for newly discovered low mass WDs given a measured WD mass and mass function.
\end{abstract}

@@

	\startdata
		\multicolumn{2}{c}{\multirow{2}{*}{Priors}} & $\mathcal{U}(0.2, 1.4)$ & $\left[\ln(40)\sigma\right]^{-1}$  & $\mathcal{U}(0, 1)$ \\
		\multicolumn{2}{c}{} & & $(0.05 < \sigma/\Msun < 2)$ & \\
		\cutinhead{Test Cases}
		\multirow{2}{*}{Test 1} & True & 0.7 & 0.2 & 0.0 \\
		 & MAP & {\bf XXX} & {\bf XXX} & {\bf XXX} \\
		 \hline
		\multirow{2}{*}{Test 2} & True & 0.7 & 0.2 & 0.1 \\
		 & MAP & {\bf XXX} & {\bf XXX} & {\bf XXX} \\
		 \hline
		\multirow{2}{*}{Test 3} & True & \nodata & \nodata & 0.1 \\
		 & MAP & {\bf XXX} & {\bf XXX} & {\bf XXX} \\
		 \cutinhead{ELM Sample}
		 & MAP & 0.73 & 0.26 & {\bf 5\%?}
	\enddata

@@

We use an ensemble Markov Chain Monte Carlo algorithm \citep{goodman10} to draw samples from the posterior distribution, $p(\mu_{\rm WD}, \sigma_{\rm WD}, f_{\rm NS} \given \bs{m}_f, \bs{M}_1)$. The algorithm is implemented in the \texttt{Python} programming language \citep{foremanmackey13} and uses an ensemble of individual ``walkers'' to naturally adapt to the geometry of the parameter-space being explored. We run the walkers for an initial, burn-in period of 1000 steps starting from randomly drawn initial conditions (sampled from the priors in Table~\ref{tbl:parameters}). We then re-initialized the walkers from their positions at the end of this run, throw out the burn-in samples, and run again for 1000 steps. We compute the autocorrelation time for each parameter to confirm they have converged.

@@

As a first test, we generate $M_2$'s by drawing from a single, truncated Gaussian with parameters given in \ref{tbl:parameters}. This test sample contains no NSs. The leftmost panel in Figure~\ref{fig:ELM_post} shows that the MAP $M_2$ distribution qualitatively matches the input distribution, while the second panel shows the posterior distribution quantitatively matches the input values shown as dashed lines. The third panel shows that although the posterior $f_{\rm NS}$ distribution is consistent with 0\%, there is a tail up to $\approx 10$\%. Nevertheless, our model closely recovers the input parameters. The fourth panel lists individual systems, ordered by $\mf$, and provides the distribution of $P_{\rm NS}$ values for each. Few systems have $P_{\rm NS}$ values higher than $\approx$5\%, consistent with our input distribution of $f_{\rm NS}=$0\%.

\subsection{Test 2: Two Gaussians (white dwarf + neutron star)} \label{sec:exp2}
For our next test, we use the same Gaussian distribution to generate companion masses for the white dwarf component as the previous example (Section~\ref{sec:exp1}), but we now add an additional neutron star component with parameters, setting $f_{\rm NS} =$ 10\%. The panels are the same as in Test 1 described above. Again, our model recovers the input values for $\mu{\rm WD}$ and $\sigma_{\rm WD}$. Importantly, the third panel shows that it also recovers $f_{\rm NS}$, however the posterior again shows a tail toward higher probabilities. The fourth panel indicates that some systems have near unity $P_{\rm NS}$ values; our model correctly identifies these as true NS. However, this accounts for only roughly half of the systems with NSs in the sample, and the other half may found with low $\mf$ due to a low inclination angle. Although our model can correctly identify the $f_{\rm NS}$ within a factor of $\approx2$, it can only identify some LMWDs as having NS companions.

% [TODO: We have also tried changing the mean and variance of the true neutron star distribution while keeping the fitting parameters fixed and found ...]

\subsection{Test 3: Uniform (white dwarf) + Gaussian (neutron star)} \label{sec:exp3}
For a final test, we generate companion masses for the white dwarf component by sampling from a uniform distribution over the range $(0.2,1.2)~\Msun$, while keeping $f_{\rm NS}=$10\%. Again, the panels are the same as in Test 1 described above. The posterior distribution in the second panel shows that the systems have difficulty converging on any one solution for $\mu_{\rm WD}$ and $\sigma_{\rm sigma}$. Interestingly, the third panel shows that despite having a non-Gaussian input distribution for $M_2$, our model still recovers $f_{\rm NS}$, approximately as accurately as in Test 2. %The rightmost panel shows that despite accurately finding $f_{\rm NS}$, our model is not as good at identifying exactly which WDs are NSs.

@@

\begin{figure*}[h!]
\begin{center}
\includegraphics[width=0.95\textwidth]{real-data.pdf}
\caption{The posteriors for our model applied to the ELM WDs. The first panel shows the MAP $M_2$ distribution (solid black) and random samples from the posterior (grey lines). The second panel shows the posterior distribution of $\mu$ and $\sigma$; The maximum a posteriori Gaussian model has $\mu_{\rm WD}= {\bf 0.74}~\Msun$ and $\sigma_{\rm WD}= {\bf 0.26}~\Msun$. The third panel the posterior distribution of $\mu$ and $f_{\rm NS}$. The last panel lists individual LMWDs in the ELM sample and shows each of their corresponding NS probabilities. The systems are ordered by increasing $\mf$. The three systems with $P_{\rm NS}=$0\% are the three eclipsing systems.}
\label{fig:ELM_post}
\end{center}
\end{figure*}

@@

The results of our model are shown in Figure \ref{fig:ELM_post}. The maximum a posteriori Gaussian model has $mu_{\rm WD} = {\bf 0.74} \Msun$ and $\sigma_{\rm WD} = {\bf 0.26}~\Msun$. While the well-defined peak in the posterior distribution demonstrates convergence of the model, the extended distribution toward larger variances could indicate that the true distribution does not match a Gaussian. However it is interesting that the best fit Gaussian for the companions to the ELM WD sample is similar to that of the population of single DA WDs in SDSS with a mean of 0.6 $\Msun$ \citep{kleinman13}. Our distribution is significantly wider ({\bf $\sigma \approx 0.26~\Msun$} compared with $\sigma \approx 0.1~\Msun$), likely due to mass transfer affecting the mass distribution of the unseen primary WDs as well.

@@

The third panel in Figure~\ref{fig:ELM_post} shows a NS fraction strongly peaked toward zero, however there is a significant tail toward higher NS probabilities. Our model does indicate an upper limit to the NS fraction of {\bf XXX\%} at the 95\% confidence level. For 55 systems with observed radial velocity variations, this corresponds to a {\bf 0-4?} systems with NS companions. These results are in agreement with independent constraints from \citet{vLeeuwen07} who set an upper limit on $f_{\rm NS}$ of 18$\pm5$\% based on the non-detection of radio emission from eight LMWD targets. With non-detections of radio emission from an expanded sample of 15 LMWD targets in the ELM WD sample, \citet{agueros09b} place a more stringent upper limit on $f_{\rm NS}$ of $10\substack{+4 \\ -2}\%$.

Since our results are still consistent with a 0\% NS fraction, it is not guaranteed that any of the LMWDs in the ELM WD sample have NS companions. However the fourth panel in Figure \ref{fig:ELM_post} indicates which ELM WDs have the highest probability of hosting NS companions. Here, we show the relation between $\mf$ and $P_{\rm NS}$ for individual systems. This panel shows that $P_{\rm NS}$ only starts to significantly deviate from zero for {\bf $\mf\approx0.6 \Msun$} Although a few systems have high probabilities, the distribution is strongly clustered near 0\%.

The two systems with the highest $P_{\rm NS}$ are J0811$+$0225 and J1741$+$6526. These systems are ideal candidates for searches for NS companions in radio and X-ray. Radio observations of these objects are in progress, however because a NS companion may be radio quiet or beamed away from us, a NS companion to these LMWDs may not be detectable in radio \citep{vLeeuwen07}. All NS give off blackbody radiation peaked in X-ray wavelengths, therefore X-ray follow-up provides a more stringent test on the potential for a NS companion \citep{agueros09a}. Previous searches for LMWDs with X-ray counterparts have all been non-detections, however we show here that all those sources have a low probability of hosting a NS companion. If J0811$+$0225 and J1741$+$6526 also lack X-ray counterparts, then even more stringent constraints can be placed on $f_{\rm NS}$. In future work, we plan to combine these results with radio and x-ray non-detections to further constrain $P_{\rm NS}$ for each system.

@@

To determine the probability of any particular LMWD of hosting a NS, we make our model posteriors publicly available on fig{\bf share} \footnote{\url{http://figshare.com/authors/Jeff\_Andrews/644768}}. We further provide a {\tt python} script that calculates $P_{\rm NS}$ and the mass distribution for a WD companion for any LMWD with a measured $M_1$ and $\mf$. This script can be applied to newly discovered LMWDs as well as those already in the ELM sample.

>>>>

\begin{abstract}
Measuring the masses of companions to single-line spectroscopic binary stars
is (in general) not possible because of the unknown orbital plane inclination angles. 
Even when the mass of the visible star can be measured, only a lower limit can be placed
on the mass of the unseen companion. However, since these inclination angles should be isotropically
distributed, for a large enough, unbiased sample, the 
companion mass distribution can be deconvolved from the distributions of observables.
In this work, we construct a hierarchical probabilistic model to infer properties of unseen 
companion stars given orbital period and projected radial velocity measurements of the primary. We use a mixture of 
two Gaussians to model the true companion mass distribution to represent 
populations of white dwarfs (WD) and neutron stars (NS). We run this model on mock data sets in order to test the
robustness of our inferences. We then apply our model to 55 of the 61 white dwarfs in 
the extremely low mass (ELM) WD survey. We find maximum a posteriori (MAP) parameters for
the WD companion population of $\mu_{\rm WD} = 0.71~\Msun$ and $\sigma_{\rm WD} = 0.26~\Msun$ 
with a NS fraction consistent with zero, $f_{\rm NS} = 0\%$. However, the marginal posterior 
distribution over NS fraction has a significant tail up to $f_{\rm NS} \approx 16\%$.
We make samples from the posterior distribution publicly available so that future observational
efforts may compute the NS probability for newly discovered low mass WDs given a measured 
WD mass and mass function.
\end{abstract}

@@

	\startdata
		\multicolumn{2}{c}{\multirow{2}{*}{Priors}} & $\mathcal{U}(0.2, 1)$ & $\left[\ln(100)\sigma\right]^{-1}$  & $\mathcal{U}(0, 1)$ \\
		\multicolumn{2}{c}{} & & $(0.02 < \sigma/\Msun < 2)$ & \\
		\cutinhead{Test Cases}
		\multirow{2}{*}{Test 1} & True & 0.7 & 0.2 & 0.0 \\
		 & MAP & {\bf 0.74} & {\bf 0.18} & {\bf 0.0} \\
		 \hline
		\multirow{2}{*}{Test 2} & True & 0.7 & 0.2 & 0.1 \\
		 & MAP & {\bf 0.76} & {\bf 0.17} & {\bf 0.12} \\
		 \hline
		\multirow{2}{*}{Test 3} & True & \nodata & \nodata & 0.1 \\
		 & MAP & {\bf 0.80} & {\bf 0.39} & {\bf 0.12} \\
		 \cutinhead{ELM Sample}
		 & MAP & 0.71 & 0.26 & {\bf 0.0?}
	\enddata

@@

We use an ensemble Markov Chain Monte Carlo algorithm \citep{goodman10} to draw samples from the posterior distribution, $p(\mu_{\rm WD}, \sigma_{\rm WD}, f_{\rm NS} \given \bs{m}_f, \bs{M}_1)$. The algorithm is implemented in the \texttt{Python} programming language \citep{foremanmackey13} and uses an ensemble of individual ``walkers'' to naturally adapt to the geometry of the parameter-space being explored. We run the walkers for an initial, burn-in period of 500 steps starting from randomly drawn initial conditions (sampled from the priors in Table~\ref{tbl:parameters}). We then re-initialized the walkers from their positions at the end of this run, throw out the burn-in samples, and run again for 1000 steps. We compute the autocorrelation time for each parameter to confirm they have converged.

@@

As a first test, we generate $M_2$'s by drawing from a single, truncated Gaussian with parameters given in Table~\ref{tbl:parameters}. This test sample contains no NSs. The leftmost panel in Figure~\ref{fig:ELM_post} shows that the MAP $M_2$ distribution (black line) qualitatively matches the input distribution (grey histogram). The second panel shows samples from the posterior distribution and the contours contain 68\% and 95\% of the samples. The input values lie cleanly within the inner contour in both projections of the posterior shown. The third panel shows that although the posterior $f_{\rm NS}$ distribution is consistent with 0\%, there is a tail up to $\sim 10$\%. Nevertheless, our model recovers the input parameters for the white dwarf companion population. The fourth panel lists individual systems, ordered by $\mf$, and shows the distributions of $P_{\rm NS}$ for each. Few systems have $P_{\rm NS}$ values higher than $\sim$5\%.

\subsection{Test 2: Two Gaussians (white dwarf + neutron star)} \label{sec:exp2}
For our next test, we use the same Gaussian distribution to generate companion masses for the white dwarf component as the previous example (Section~\ref{sec:exp1}), but we now add an additional neutron star component with $f_{\rm NS} =$ 10\% and parameters described in Section~\ref{sec:experiments}. The panels are the same as in Test 1 described above. Again, our model recovers the input values for $\mu{\rm WD}$ and $\sigma_{\rm WD}$. Importantly, the third panel shows that it also recovers $f_{\rm NS}$, however the posterior again shows a tail toward higher probabilities. The fourth panel indicates that some systems have near unity $P_{\rm NS}$ values; our model correctly identifies these as true NS. However, this accounts for only roughly half of the systems with NSs in the simulated sample; the other half have low $\mf$'s due to low inclination angles.

\subsection{Test 3: Uniform (white dwarf) + Gaussian (neutron star)} \label{sec:exp3}
For a final test, we generate companion masses for the white dwarf component by sampling from a uniform distribution over the range $(0.2,1.2)~\Msun$, while keeping $f_{\rm NS}=$10\%. The panels are the same as in Test 1 described above. The posterior distribution in the second panel shows that the systems have difficulty converging on any one solution for $\mu_{\rm WD}$ and $\sigma_{\rm sigma}$. Interestingly, the third panel shows that despite having a non-Gaussian input distribution for $M_2$, our model still recovers $f_{\rm NS}$ approximately as accurately as in Test 2. %The rightmost panel shows that despite accurately finding $f_{\rm NS}$, our model is not as good at identifying exactly which WDs are NSs.

@@

\begin{figure*}[h!]
\begin{center}
\includegraphics[width=0.95\textwidth]{real-data.pdf}
\caption{The posteriors for our model applied to the ELM WDs. The first panel shows the MAP $M_2$ distribution (solid black) and random samples from the posterior (grey lines). The second panel shows the posterior distribution of $\mu$ and $\sigma$; The maximum a posteriori Gaussian model has $\mu_{\rm WD}= {\bf 0.71}~\Msun$ and $\sigma_{\rm WD}= {\bf 0.26}~\Msun$. The third panel the posterior distribution of $\mu$ and $f_{\rm NS}$. The last panel lists individual LMWDs in the ELM sample and shows each of their corresponding NS probabilities. The systems are ordered by increasing $\mf$. The three systems with $P_{\rm NS}=$0\% are the three eclipsing systems.}
\label{fig:ELM_post}
\end{center}
\end{figure*}

@@

The results of using our model on the ELM sample are shown in Figure \ref{fig:ELM_post}. The maximum a posteriori model has parameters $\mu_{\rm WD} = {\bf 0.71} \Msun$, $\sigma_{\rm WD} = {\bf 0.26}~\Msun$, and $f_{\rm NS} = 0\%$. The marginal posterior over $\mu_{\rm WD}$ and $\sigma_{\rm WD}$ has a tail towards larger variances, which could indicate that the true WD distribution is not exactly Gaussian. However, it is interesting that the best fit Gaussian for the companions to the ELM WD sample is similar to that of the population of single DA WDs in SDSS with a mean of 0.6 $\Msun$ \citep{kleinman13}. Our distribution is significantly wider ({\bf $\sigma \approx 0.26~\Msun$} compared with $\sigma \approx 0.1~\Msun$), possibly due to mass transfer affecting the mass distribution of the unseen primary WDs as well.

@@

The third panel in Figure~\ref{fig:ELM_post} shows a NS fraction strongly peaked toward zero, however there is a significant tail toward higher NS probabilities. Our model indicates an upper limit to the NS fraction of {\bf 25\%} at the 95\% confidence level. For 55 systems with observed radial velocity variations, this corresponds to a {\bf 0-4?} systems with NS companions. {\bf [Jeff: TODO -- this probably needs to be updated?]} These results are in agreement with independent constraints from \citet{vLeeuwen07} who set an upper limit on $f_{\rm NS}$ of 18$\pm5$\% based on the non-detection of radio emission from eight LMWD targets. With non-detections of radio emission from an expanded sample of 15 LMWD targets in the ELM WD sample, \citet{agueros09b} place a more stringent upper limit on $f_{\rm NS}$ of $10\substack{+4 \\ -2}\%$.

%Since our results are still consistent with a 0\% NS fraction, it is not guaranteed that any of the LMWDs in the ELM WD sample have NS companions. However the fourth panel in Figure \ref{fig:ELM_post} indicates which ELM WDs have the highest probability of hosting NS companions. Here, we show the distributions of $P_{\rm NS}$ for each system, ordered by $\mf$. This panel shows that $P_{\rm NS}$ only starts to significantly deviate from zero for {\bf $\mf\approx0.6 \Msun$} Although a few systems have high probabilities, the distribution is strongly clustered near 0\%.

The two systems with the highest $P_{\rm NS}$ are J0811$+$0225 and J1741$+$6526. These systems are ideal candidates for searches for NS companions in radio and X-ray. Radio observations of these objects are in progress, however because a NS companion may be radio quiet or beamed away from us, a NS companion to these LMWDs may not be detectable in radio \citep{vLeeuwen07}. All NS give off blackbody radiation peaked in X-ray wavelengths, therefore X-ray follow-up provides a more stringent test on the potential for a NS companion \citep{agueros09a}. Previous searches for LMWDs with X-ray counterparts have all been non-detections, however we show here that all those sources have a low probability of hosting a NS companion. If J0811$+$0225 and J1741$+$6526 also lack X-ray counterparts, then even more stringent constraints can be placed on $f_{\rm NS}$. % In future work, we plan to combine these results with radio and x-ray non-detections to further constrain $P_{\rm NS}$ for each system.

@@

{\bf Jeff: change your mind, or is this old?}
To determine the probability of any particular LMWD of hosting a NS, we make our model posteriors publicly available on fig{\bf share} \footnote{\url{http://figshare.com/authors/Jeff\_Andrews/644768}}. We further provide a {\tt python} script that calculates $P_{\rm NS}$ and the mass distribution for a WD companion for any LMWD with a measured $M_1$ and $\mf$. This script can be applied to newly discovered LMWDs as well as those already in the ELM sample.
